\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}

\newcommand{\lean}[1]{\texttt{#1}}
\newcommand{\vv}[1]{\textsf{\small #1}}

\title{Reconciling a multi-voice corpus by set-theoretic estimation:\\
the ``cause of the tides'' run\\
\large Haiku extracted the frames; Lean machine-checked the verdict}
\author{thejeb project note}
\date{2026-07-24}

\begin{document}
\maketitle

\begin{abstract}
We ran the empirical pipeline of the companion plan end to end. Seven
public-domain natural philosophers discussing one topic --- the cause of
the ocean tides --- were extracted into a structured frame per author by
\textbf{Claude Haiku 4.5} (open pass), unified into a canonical schema by
\textbf{Claude Sonnet}, and re-extracted against that schema by Haiku; the
resulting frames were encoded as STE property sets and the invariants of
the disagreement were \textbf{machine-checked in Lean} against
\lean{Ste.Basic} (Combettes 1993). The verdict: the corpus is globally
\emph{inconsistent}; the obstruction is carried by the moon's role, which
splits three ways (attraction / pressure / rejected); and the
disagreement is \emph{pervasive} rather than localized to one outlier ---
even the five-author attraction camp is internally inconsistent on the
mechanism. This corrected the plan's armchair prediction of a single
outlier (Galileo): the machine found \emph{two} dissenters and a deeper
coupling, which is the point of running it.
\end{abstract}

\section{The run}

The pipeline did not fix the schema up front. Three passes (prompts
committed in \lean{scripts/tides/}, outputs in
\lean{sources/tides/extraction/}):

\begin{enumerate}
\item \textbf{Haiku, open extraction.} Haiku read the seven passages and
  proposed its own $(\text{variable},\text{value},\text{quote})$ claims,
  over-splitting into $\approx 25$ fine-grained variables
  (\lean{pass1.json}).
\item \textbf{Sonnet, schema unification.} Sonnet merged the synonymous
  variables into a canonical \textbf{7-variable} schema with closed
  domains and a synonym map (\lean{schema.json}): \vv{primary\_cause},
  \vv{moon\_role} $\in\{$attraction, pressure, rejected$\}$,
  \vv{mechanism} (5 values), \vv{sun\_role}, \vv{spring\_neap},
  \vv{earth\_motion\_required}, \vv{quantitative}.
\item \textbf{Haiku, schema-constrained re-extraction.} Haiku re-extracted
  every author against the canonical schema, emitting one assignment per
  author (value or \vv{silent}) with a licensing quote
  (\lean{frames.json}).
\end{enumerate}

The extraction ran on Haiku 4.5 / Sonnet through the agent harness; the
prompts are reusable and reproduce the frames.

\section{The extracted frames}

The canonical frames (\vv{$\cdot$} $=$ silent):

\begin{center}\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}l l l l l l@{}}
\toprule
Author & primary\_cause & moon\_role & mechanism & sun\_role & quant. \\
\midrule
Pliny     & sun+moon    & \textbf{attraction} & direct\_corr.   & reinforces & $\cdot$ \\
Bede      & moon        & \textbf{attraction} & direct\_corr.   & $\cdot$    & $\cdot$ \\
Kepler    & moon        & \textbf{attraction} & gravitational   & $\cdot$    & $\cdot$ \\
Galileo   & earth\_motion & \textbf{rejected} & basin\_motion   & $\cdot$    & $\cdot$ \\
Descartes & sun+moon    & \textbf{pressure}   & vortex\_pressure& reinforces & $\cdot$ \\
Newton    & sun+moon    & \textbf{attraction} & gravitational   & weaker     & math. \\
Laplace   & sun+moon    & \textbf{attraction} & dyn.\ oscill.   & $\cdot$    & math. \\
\bottomrule
\end{tabular}
\end{center}

\section{The machine-checked invariants}

Encoded in \lean{Ste.TidesCorpus}: each author $a$ is a property set
$\lean{constraint}\,a = \{f \mid f \text{ agrees with } a\text{'s frame
where } a \text{ speaks}\}$, and the STE feasibility set is
$\lean{feasibilitySet constraint} = \bigcap_a \lean{constraint}\,a$. All
of the following are proved (sorry-free):

\begin{itemize}
\item \textbf{Global inconsistency} (\lean{feasible\_eq\_empty}):
  $\lean{feasibilitySet constraint} = \varnothing$. No single reading of
  the tides satisfies all seven voices.
\item \textbf{The three-way \vv{moon\_role} split gives three minimal
  cores}: any two authors in different camps already have disjoint
  property sets ---
  \lean{kepler\_galileo\_core} (attraction vs.\ rejected),
  \lean{kepler\_descartes\_core} (attraction vs.\ pressure),
  \lean{galileo\_descartes\_core} (rejected vs.\ pressure).
\item \textbf{The disagreement is pervasive}
  (\lean{pliny\_kepler\_mechanism\_core}): even inside the five-author
  attraction camp, Pliny (direct correspondence) and Kepler
  (gravitational) are a minimal conflicting core on \vv{mechanism}. So
  dropping the two \vv{moon\_role} dissenters does \emph{not} restore
  consistency.
\item \textbf{Positive structure on the most-contested variable}
  (\lean{attraction\_camp\_moon\_consistent},
  \lean{descartes\_dissents}, \lean{galileo\_dissents}): on
  \vv{moon\_role} alone, the five attraction authors share a common
  reading, and exactly two voices dissent --- Descartes (pressure) and
  Galileo (rejected).
\item \textbf{The Combettes corollary} (\lean{no\_fair\_reading}): because
  the formulation is inconsistent, every candidate estimand is unfair to
  some voice --- the set-theoretic detection-of-invalid-information
  principle, instantiated on the corpus.
\end{itemize}

\section{What the machine corrected}

The companion plan predicted, from historical intuition, a \emph{single}
famous outlier (Galileo) and a nonempty consensus once he is dropped. The
run refuted both halves:

\begin{enumerate}
\item There are \textbf{two} dissenters on the moon's role, not one:
  Galileo (rejected) \emph{and} Descartes (pressure). The
  \vv{moon\_role} variable is a genuine three-way split.
\item Dropping the dissenters does \textbf{not} yield a consensus: the
  attraction camp is itself split on \vv{mechanism} (direct
  correspondence vs.\ gravitational attraction vs.\ dynamical
  oscillation). The coupling is layered and pervasive.
\end{enumerate}

This is exactly the value of a machine-checked empirical instance: the
verdict is derived from the extracted frames, not from a plausible story,
and it overturned the plausible story.

\section{The boundary of trust}

The Lean layer certifies the aggregation logic, not the reading of the
text. Every asserted value in \lean{frames.json} carries the source
sentence that licenses it, so each frame is auditable against the
passage; but a mis-extraction would yield a wrong-but-machine-checked
verdict. The invariants above are conditional on the frames. The corpus
passages are curated public-domain representative statements (provenance
in \lean{refs.bib}); pinning exact critical editions and expanding beyond
seven voices is future corpus work.

\section{The computable layer: invariants the kernel computes}

The qualitative module above proves its verdict with hand-written
\lean{Set} arguments. Plan phase P2 asked for the \emph{computable}
refinement: make the aggregation executable and prove the computable layer
agrees with the verified \lean{Set}-based spec. This is now done, in two
modules.

\lean{Ste.FiniteInstance} proves the general structural collapse. For a
finite frame corpus --- each author fixing a partial assignment
$\lean{frame}\,a : V \to \mathtt{Option}\,W$ (silence $=$ \lean{none}) ---
joint satisfiability is \emph{not} a genuinely set-level question. It is
equivalent to a \textbf{decidable, pairwise} one: the corpus is satisfiable
iff no two authors ever assign the same variable two different values. The
key theorem
\[
  \lean{feasibilitySet}\,(\lean{frameConstraint}\;\lean{frame}) = \varnothing
  \;\Longleftrightarrow\; \neg\,\lean{Consistent}\;\lean{frame}
\]
(\lean{feasibilitySet\_eq\_empty\_iff}) ties the verified \lean{Ste.Basic}
feasibility set (left) to a \lean{Decidable} predicate (right), via a
canonical witness \lean{frameWitness}. It also defines the quantitative
\lean{disagreementDegree}\,$v$ (the number of distinct values variable $v$
receives) with the characterization that a variable is unanimous-where-
spoken iff its degree is $\le 1$ (\lean{agree\_iff\_degree\_le\_one}).

\lean{Ste.TidesComputable} instantiates this on the actual frames ---
$\lean{constraint} = \lean{frameConstraint}\;\lean{frame}$ holds by
\lean{rfl}, so every general theorem specializes with no glue --- and lets
the \textbf{Lean kernel compute} the invariants by \lean{decide}:

\begin{itemize}
\item \lean{tides\_inconsistent}: $\neg\,\lean{Consistent}\;\lean{frame}$,
  decided by the kernel; and \lean{feasible\_eq\_empty\_computed} feeds it
  through the bridge to re-derive $\lean{feasibilitySet constraint} =
  \varnothing$ --- the \emph{same} proposition as the hand-proved
  \lean{feasible\_eq\_empty}, now obtained by computation. The computable
  layer is checked to agree with the verified spec.
\item The \textbf{disagreement degrees}, each by \lean{decide}:
  \vv{primary\_cause}$=3$, \vv{moon\_role}$=3$, \vv{mechanism}$=5$,
  \vv{sun\_role}$=2$, and \vv{spring\_neap} $=$ \vv{earth\_motion} $=$
  \vv{quantitative}$=1$. The two headline conflict sites are exactly the
  three-way \vv{moon\_role} split and the five-way \vv{mechanism} split.
\item \lean{conflictVars\_card}$=4$ and \lean{conflictVars\_eq}: precisely
  four variables (\vv{primary\_cause}, \vv{moon\_role}, \vv{mechanism},
  \vv{sun\_role}) carry the disagreement --- computed, not asserted.
\item \lean{springNeap\_unanimous}: obtained \emph{through} the degree
  characterization, a positive fact --- every author who mentions
  spring/neap dependence agrees.
\end{itemize}

The computation uses the Lean \textbf{kernel} (\lean{decide}), not
\lean{native\_decide}, so the compiler stays out of the trusted base.

\section{Verified now vs.\ next}

Mechanized: the feasibility verdict (empty), the minimal conflicting
cores, the three-way split, the pervasive-coupling core, the
single-variable consensus/dissent structure, \emph{and} --- now computably
--- the decidable feasibility bridge, the per-variable disagreement
degrees, and the four conflict sites, all tied to the verified
\lean{feasibilitySet}. Not yet mechanized for this instance: the remaining
\emph{representation-theoretic} invariants --- the concept lattice of
coherent voice-clusters (\lean{Ste.HyperFrame}), the rectangle-cover
complexity $\rho$ and \v{C}ech obstruction magnitude
(\lean{Ste.RepresentationBounds}, \lean{Ste.CechObstruction}), and the
treewidth of the relation graph (\lean{Ste.Treedecomp}) run on this
instance. The feasibility verdict and its quantitative disagreement
profile are complete and kernel-checked.

\bibliographystyle{plainnat}
\bibliography{../../refs}

\end{document}
