\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}

\newcommand{\lean}[1]{\texttt{#1}}
\newcommand{\vv}[1]{\textsf{\small #1}}

\title{Set-theoretic estimation at scale: 86 documents through the\\
chunking pipeline --- The Federalist Papers\\
\large Haiku/Sonnet extracted the frames; Lean ran the proved algorithm}
\author{thejeb project note}
\date{2026-07-24}

\begin{document}
\maketitle

\begin{abstract}
We scale the empirical STE pipeline from seven short passages (the tides
note) to a large corpus of long documents: the \textbf{86 essays of The
Federalist Papers}. Each essay is one STE ``voice.'' A general,
topic-parameterized pipeline \emph{chunks} every document, extracts open
frames per chunk with \textbf{Claude Haiku}, unifies a canonical
10-variable schema with \textbf{Claude Sonnet}, and re-extracts one frame
per document against that schema; the 86 frames are then scored by the
\emph{verified} computable STE layer (\lean{Ste.FiniteInstance}) running in
a Lean executable that reads the frames at runtime. The verdict, computed
by the proved algorithm: the corpus is \emph{inconsistent}, but the
inconsistency is \emph{narrow} --- 7 of 10 variables are unanimous where
spoken (a large consistent core), and exactly \textbf{3 conflict sites}
carry the whole obstruction, headlined by a genuine three-way split on how
Publius characterizes the proposed government. The machine recovered the
real, well-documented internal tension of The Federalist.
\end{abstract}

\section{The corpus and the pipeline}

The corpus is the 86 essays of The Federalist \citep{hamilton1788federalist}
(85 numbers; No.~70 appears twice in the source edition), split from Project
Gutenberg eBook \#18 into one public-domain file per essay with a provenance
header. The median essay is $\approx 13$k characters (max $\approx 34$k) ---
long enough that per-document \emph{chunking} is required, not optional.

The pipeline (\lean{scripts/corpus/}) is the tides pipeline generalized to
any corpus and topic, with chunking and parallel model calls:

\begin{enumerate}
\item \textbf{Pass 1 --- Haiku, open, per chunk.} Each document is split into
  $\approx 9$k-char chunks on paragraph boundaries; each chunk is extracted
  independently into $(\text{variable},\text{value},\text{quote})$ claims,
  unioned per document. $86$ documents $\rightarrow$ \textbf{189 chunks}.
\item \textbf{Pass 2 --- Sonnet, unify, once.} The 189 chunks proposed
  \textbf{2{,}510 distinct} $(\text{variable},\text{value})$ pairs; Sonnet
  merged them into a canonical \textbf{10-variable} schema with closed
  domains and a synonym map.
\item \textbf{Pass 3 --- Haiku, schema-constrained, per document.} Each of
  the 86 documents is re-extracted against the canonical schema into one
  frame (a domain value or \vv{silent} per variable, with a licensing
  quote).
\end{enumerate}

The run made $\approx 276$ model calls ($189+1+86$) in parallel
(\lean{--jobs 8}) in roughly half an hour. The output \lean{frames.json} has
exactly the shape the verified Lean executable reads.

\section{The verdict --- the proved algorithm on runtime data}

\lean{lake exe tides frames.json} reads the 86 frames and evaluates the
\emph{verified} \lean{STE.Consistent} and \lean{STE.disagreementDegree} of
\lean{Ste.FiniteInstance} --- the definitions proved (in that module) to
characterize the \lean{Set}-level feasibility of \lean{Ste.Basic}
(\lean{feasibilitySet\_eq\_empty\_iff}). No \lean{native\_decide}; the
aggregation logic is the proved one, run on the loaded data.

\begin{center}\small
\begin{tabular}{@{}l c l@{}}
\toprule
Variable & degree & reading \\
\midrule
ratification\_stance        & 1 & unanimous (supports ratification) \\
union\_vs\_disunion          & 1 & unanimous (single union essential) \\
standing\_army\_restriction  & 1 & unanimous \\
executive\_structure        & 1 & unanimous (single executive) \\
judicial\_power             & 1 & unanimous (safe, essential check) \\
bill\_of\_rights\_necessity  & 1 & unanimous \\
extended\_republic\_theory   & 1 & unanimous (extended republic superior) \\
\textbf{federal\_power\_scope}     & \textbf{2} & 32 extensive vs.\ 13 strictly-limited \\
\textbf{senate\_equal\_suffrage}   & \textbf{2} & 5 justified vs.\ 1 unjust \\
\textbf{government\_character}     & \textbf{3} & mixed / federal / national \\
\bottomrule
\end{tabular}
\end{center}

\begin{itemize}
\item \textbf{Inconsistent} (\lean{STE.Consistent} $=$ \textsf{false}): the
  feasibility set is empty --- 86 essays by one pseudonymous author do not
  admit a single joint reading.
\item \textbf{But narrowly so.} \textbf{7 of 10 variables are unanimous
  where spoken} (disagreement degree $\le 1$): ratification, union, the
  standing army, the executive, the judiciary, the bill of rights, and the
  extended republic. That is a large \emph{consistent core} --- the shared
  Federalist program.
\item \textbf{Three conflict sites} carry the whole obstruction. The
  headline is \vv{government\_character}, a genuine \emph{three-way} split
  (degree 3): \vv{mixed\_composite} (25 essays, incl.\ Federalist~39's
  ``neither a national nor a federal Constitution, but a composition of
  both''), \vv{federal\_confederacy} (11 essays, e.g.\ Federalist~45's
  ``powers delegated\dots few and defined''), and
  \vv{national\_consolidated} (Hamilton's Federalist~12/15). The other two
  are \vv{federal\_power\_scope} (extensive vs.\ strictly limited) and
  \vv{senate\_equal\_suffrage}.
\end{itemize}

The pipeline recovered, from raw text and by a proved decision procedure,
the real internal tension long noted by readers of The Federalist: Publius
describes the proposed government three different ways depending on
rhetorical need, and the essays differ on the reach of federal power.

\section{Contrast with the tides}

\begin{center}\small
\begin{tabular}{@{}l l l@{}}
\toprule
 & Tides (7 voices) & Federalist (86 voices) \\
\midrule
Consistent? & no & no \\
Structure & pervasive, multi-camp & consistent core $+$ 3 conflict sites \\
Worst variable & mechanism (degree 5) & government\_character (degree 3) \\
Reading & rival theories & one program, few internal tensions \\
\bottomrule
\end{tabular}
\end{center}

Same verified algorithm; two qualitatively different disagreement shapes.
That the verdict is \emph{computed} rather than asserted is exactly what
lets it distinguish them.

\section{The boundary of trust}

The Lean layer certifies the aggregation, not the reading of the text. Each
asserted value in \lean{frames.json} carries the source sentence that
licenses it, so every cell is auditable against the essay; the three
conflict sites were spot-checked against quotes and are faithful. A
mis-extraction would move a count, but the verdict itself --- feasibility
and the disagreement degrees --- is the output of the proved algorithm on
the frames, not a plausible story about them.

\bibliographystyle{plainnat}
\bibliography{../../refs}

\end{document}
