\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}

\newcommand{\lean}[1]{\texttt{#1}}
\newcommand{\feas}{\ensuremath{\mathrm{feas}}}

\title{An empirical instance of set-theoretic estimation:\\
mechanized reconciliation of a multi-voice corpus\\
\large A plan --- Haiku extracts the frames, Lean verifies the inference}
\author{thejeb project note}
\date{2026-07-24}

\begin{document}
\maketitle

\begin{abstract}
This note fixes a concrete empirical target for the STE/frames machinery
and lays out the pipeline. The task: acquire a corpus of several authors
discussing one topic and its internal relations, extract each author's
claims into a structured \emph{frame} using a Claude Haiku model, hand the
frames to the verified Lean core, and read off the machine-checked
invariants (feasible set, concept lattice, obstruction, rectangle-cover
complexity, treewidth). The topic chosen is \textbf{the cause of the ocean
tides}, with a diachronic corpus of seven public-domain natural
philosophers spanning antiquity to the early nineteenth century. The
architecture keeps the fuzzy natural-language front-end (Haiku) strictly
separate from the machine-checked inference core (Lean), with an explicit
extraction contract marking the boundary of trust.
\end{abstract}

\section{Why this topic}

The demonstration is only compelling if the disagreement in the corpus is
\emph{structured}: partly reconcilable, partly irreducibly coupled, with a
famous outlier. The cause of the tides is the ideal first instance.

\begin{itemize}
\item \textbf{Genuine, structured disagreement.} From antiquity the tides
  were tied to the moon (\citealp{pliny77naturalis}; Seleucus of
  Seleucia; \citealp{bede725temporum}). Kepler made lunar
  \emph{attraction} the cause (\citealp{kepler1609astronomia}); Descartes
  replaced attraction with the \emph{pressure} of a celestial vortex
  (\citealp{descartes1644principia}); Newton unified sun and moon under
  \emph{universal gravitation} (\citealp{newton1687principia}); Laplace
  gave the dynamical theory (\citealp{laplace1799mecanique}). And the
  famous outlier: Galileo rejected the moon as an ``occult'' cause and
  derived the tides from the Earth's combined rotation and revolution
  (\citealp{galileo1632dialogue}, Fourth Day).
\item \textbf{The coupling is real.} Galileo's frame \emph{requires} the
  Earth to move and \emph{denies} the moon a role, which contradicts every
  other voice's affirmation of the lunar correlation. This is exactly a
  locally-coherent-but-globally-impossible configuration --- a nonzero
  Čech obstruction is expected, with Galileo an isolated concept in the
  lattice.
\item \textbf{Public-domain, bounded, legible.} Every source text is out
  of copyright and obtainable; the topic decomposes into a handful of
  discrete sub-claims with small domains; and the meaning of each invariant
  is transparent to a reader.
\end{itemize}

\section{The corpus}

Seven voices, assembled into \lean{sources/tides/} with per-item
provenance (author, work, date, edition/translation, URL, license).
Exact editions and excerpt ranges are pinned in Stage~0.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Voice & Source & Position on the cause \\
\midrule
Pliny the Elder & \emph{Naturalis Historia} II (c.~77) & moon+sun correlation \\
Bede & \emph{De temporum ratione} (725) & lunar timing \\
Kepler & \emph{Astronomia Nova} (1609) & lunar attraction \\
Galileo & \emph{Dialogue}, Fourth Day (1632) & Earth's motion; \emph{not} the moon \\
Descartes & \emph{Principia Philosophiae} (1644) & vortex pressure \\
Newton & \emph{Principia} III (1687) & universal gravitation (sun+moon) \\
Laplace & \emph{M\'ecanique c\'eleste} (1799) & dynamical (gravitational) \\
\bottomrule
\end{tabular}
\end{center}

\section{The topic's relations: the variable schema}

The ``topic and its relations'' becomes a fixed set of variables $V$ with
small per-variable domains $A_v$. Each author's frame is a partial
assignment (silence $=$ no constraint on that variable). A candidate
schema:

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Variable & Domain \\
\midrule
\lean{primary\_cause} & \{moon, sun, moon+sun, earth\_motion, vortex, other\} \\
\lean{moon\_role} & \{attraction, pressure, none\} \\
\lean{sun\_role} & \{contributes, none\} \\
\lean{mechanism} & \{gravitation, aetherial\_pressure, kinematic, occult/unspecified\} \\
\lean{requires\_earth\_motion} & \{yes, no\} \\
\lean{spring\_neap\_phase\_link} & \{affirmed, denied, silent\} \\
\lean{two\_tides\_per\_day} & \{explained, unaddressed\} \\
\lean{quantitative} & \{yes, no\} \\
\bottomrule
\end{tabular}
\end{center}

The variables are \emph{coupled} --- e.g.\ \lean{mechanism}$=$gravitation
entails \lean{moon\_role}$=$attraction and \lean{sun\_role}$=$contributes;
Galileo's \lean{requires\_earth\_motion}$=$yes with \lean{moon\_role}$=$none
conflicts with \lean{spring\_neap\_phase\_link}$=$affirmed. These edges give
the relation graph a genuine, nontrivial treewidth --- the quantity the
tractability theory is about.

\section{Architecture: Haiku to the frames, Lean from the frames on}

\begin{verbatim}
  [Claude Haiku]  corpus text --> per-author structured frame (strict JSON)
        |         claude-haiku-4-5, output_config.format = json_schema
        |         (fuzzy, unverified: all real-world risk lives here)
        v
  [handoff]       V, per-variable domains A_v, per-author constraint S_i
        v
  [Lean core]     computable STE instance + adaptive-consistency solver,
        |         proven to refine the verified Set-based spec
        v         outputs, each with a theorem behind it:
   * feasibilitySet = globally consistent readings (or empty)
   * concept lattice = coherent voice-clusters (Ste.HyperFrame)
   * cechObstruction / rho = irreducibility of the disagreement
   * inducedTreewidth = was reconciliation tractable
   * feasible  -> extracted consensus reading (backtrack-free)
   * infeasible-> minimal conflicting cores
\end{verbatim}

\subsection*{Stage 0 --- corpus assembly}
Collect the seven excerpts on the tides into \lean{sources/tides/} with a
provenance manifest; cite every source in \lean{refs.bib} (done: entries
\lean{galileo1632dialogue}, \lean{kepler1609astronomia},
\lean{descartes1644principia}, \lean{newton1687principia},
\lean{laplace1799mecanique}, \lean{pliny77naturalis},
\lean{bede725temporum}).

\subsection*{Stage 1 --- frame extraction (Claude Haiku)}
For each author-excerpt, one Haiku call with a \emph{strict output schema}
returns the frame. Concretely: \lean{claude-haiku-4-5} via the Messages
API with \lean{output\_config.format} a \lean{json\_schema} whose
properties are exactly the variables of \S4, each an \lean{enum} over its
domain plus a \lean{"silent"} option; a \lean{quote} field per assigned
variable carrying the sentence that licenses it (auditability). Extraction
is cheap ($\$1/\$5$ per Mtok; a seven-voice corpus is a handful of calls),
and corpus-scale runs use the Batches API at half price. A second
normalization pass maps free-text values onto the fixed domain vocabulary.
The model does constrained \emph{extraction}, never inference --- that is
the core's job.

\subsection*{Stage 2 --- handoff}
The JSON frames become the STE instance: the variable set $V$, the
per-variable domains $A_v$ (the union of asserted values), and one
constraint $S_i$ per author (the partial assignment it asserts). This is
the exact input shape of the bucket network of \lean{Ste.AdaptiveSolver}
(\lean{bucketNetworkList}), whose joint constraint is literally
\lean{adaptiveSet}.

\subsection*{Stage 3 --- Lean inference (verified core)}
A \emph{computable refinement}: a \lean{Finset}/decidable mirror of the
$\mathrm{Set}$-based constraint model plus the adaptive-consistency solver,
proven to agree with the verified spec, so the aggregation
\emph{executes} (\lean{\#eval}/compiled) rather than being re-implemented
on faith. It reports: the feasible set (globally consistent readings); the
concept lattice of coherent voice-clusters
(\lean{partialFeasibilitySet\_eq\_lowerPolar}, \lean{Ste.HyperFrame}); the
Čech obstruction and rectangle-cover complexity $\rho$ measuring how
irreducibly the authors disagree (\lean{Ste.CechObstruction},
\lean{Ste.RepresentationBounds}); the induced treewidth of the relation
graph (\lean{Ste.Treedecomp}); and, when feasible, an extracted consensus
reading via backtrack-free extraction
(\lean{adaptiveNetwork\_solved\_by\_elimination}) --- when infeasible, the
minimal conflicting cores.

\subsection*{Stage 4 --- writeup}
Report the invariants and interpret them, with the honest limits below in
plain view.

\section{The extraction contract and its limits}

The Lean core certifies the \emph{aggregation logic}, not the
\emph{reading of the text}. The soundness of the whole is bounded by the
natural-language$\to$constraint map of Stage~1: garbage in, \emph{verified}
garbage out. Two honest boundaries, to be stated in any writeup:

\begin{enumerate}
\item \textbf{Extraction fidelity.} Haiku's frame for each author is an
  interpretation; the \lean{quote} field makes every assigned value
  auditable against the source sentence, but a wrong reading produces a
  wrong-but-machine-checked conclusion. The invariants are conditional on
  the frames.
\item \textbf{The computable mirror is real work.} The current modules are
  specifications ($\mathrm{Set}$-valued, partly \lean{noncomputable}); the
  decidable refinement and its agreement proof are a bounded but genuine
  sub-project (Stage~3).
\end{enumerate}

\section{The demonstration payoff (predicted)}

If the frames come out as the history suggests, the machine-checked
invariants should show:
\begin{itemize}
\item a \textbf{nonzero Čech obstruction} --- no single global reading
  satisfies all seven voices, because Galileo's Earth-motion frame denies
  the lunar cause the other six require;
\item Galileo as an \textbf{isolated concept} in the lattice, with the
  ancient/lunar voices, the attraction voices (Kepler, Newton, Laplace),
  and the vortex voice (Descartes) forming the coherent clusters;
\item a \textbf{small $\rho$} --- the disagreement is covered by a few
  coherent camps, not $2^{|V|}$ arbitrary readings;
\item a \textbf{low treewidth} on the relation graph --- the topic is
  cheap to reconcile once the outlier is removed, and dropping Galileo's
  constraint yields a nonempty feasible set (the modern synthesis).
\end{itemize}
That quantitative verdict --- \emph{this corpus's disagreement is coupled
through exactly one variable, with obstruction localized to one voice} ---
machine-certified, is precisely what a pure-embedding or summarization
pipeline cannot deliver.

\section{Phasing}

\begin{enumerate}
\item[P0.] Stage~0 corpus + provenance manifest; bib done.
\item[P1.] Fix the variable schema; write and validate the Haiku
  extraction prompt + JSON schema on the seven excerpts.
\item[P2.] Build the computable Lean core (\lean{Finset} refinement +
  agreement proof); CI-green.
\item[P3.] Wire Stage~2 handoff; run Stage~3; produce the invariant table.
\item[P4.] Writeup with the honest-limits section; build the PDF.
\end{enumerate}

\bibliographystyle{plainnat}
\bibliography{../../refs}

\end{document}
