\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{listings}
\usepackage[numbers,sort&compress]{natbib}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}

\lstset{basicstyle=\ttfamily\small,columns=fullflexible,breaklines=true,
  frame=single,framesep=4pt,xleftmargin=4pt}

\title{A Machine-Checked Core for Set Theoretic Estimation}
\author{Project \texttt{thejeb}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We report a first, machine-checked formalization in Lean~4 / Mathlib of
the core of set theoretic estimation (STE) as founded by
\citet{combettes1993} and recast over topological spaces for decryption
by \citet{carlson2012}. The formalization covers the feasibility set and
its lattice-theoretic properties, the consistent / fair / ideal
taxonomy, information monotonicity, an inconsistency-based
invalid-information detector, closedness and compactness-based existence
of estimates on topological spaces, convexity of the feasibility set,
and a skeleton of Carlson's STE-decryption reduction. Every theorem
stated here is verified by continuous integration on each push; nothing
is asserted proven otherwise.
\end{abstract}

\section{Introduction}
Set theoretic estimation abandons the search for an optimum in favor of
\emph{feasibility}: an estimate is acceptable exactly when it is
consistent with every available piece of information
\citep{combettes1993}. This shifts the mathematical content from analysis
of an objective function to the set algebra and topology of a family of
constraint sets and their intersection. That content is a natural target
for a proof assistant. We mechanize it in Lean~4 on Mathlib (pinned
\texttt{v4.32.0}), adopting a strict discipline: a statement counts as
established only when it type-checks in our library under the project's
CI. This paper is the human-readable companion to that library; every
displayed result names its Lean identifier.

\section{The formal model}
Let $\Xi$ be the solution space and $I$ an index set of information
pieces. A \emph{property set family} is $S:I\to\mathcal P(\Xi)$; $S_i$
holds the estimates consistent with information $i$. We mechanize the
crisp case (Combettes's membership function $\Psi_i$ being
$\{0,1\}$-valued, so $S_i$ is a subset).

\begin{definition}[feasibility set; \texttt{STE.feasibilitySet}]
$S \;=\; \bigcap_{i\in I} S_i$ (Combettes Eq.~4). A point of $S$ is a
\emph{set theoretic estimate}.
\end{definition}

\begin{lstlisting}
def feasibilitySet (S : I -> Set Xi) : Set Xi := iInter i, S i
theorem mem_feasibilitySet : a in feasibilitySet S <-> forall i, a in S i
\end{lstlisting}

\section{Results (all machine-checked)}

\paragraph{Refinement and monotonicity.}
The feasibility set refines each constraint
(\texttt{feasibilitySet\_subset}: $S\subseteq S_j$). Writing
$S_J=\bigcap_{i\in J}S_i$ for the \emph{partial} feasibility set,
enforcing more information shrinks it:
\begin{proposition}[\texttt{partialFeasibilitySet\_antitone}]
$J\subseteq K \implies S_K \subseteq S_J$.
\end{proposition}
\noindent This is the STE reading of ``more information never enlarges
the acceptable set,'' and it descends to Carlson's key search as
\texttt{feasibleKeys\_antitone}.

\paragraph{Consistency, fairness, invalid information.}
Call the family \emph{fair} for $h$ when $h\in S_i$ for all $i$
(\texttt{STE.Fair}); this is exactly $h\in S$
(\texttt{fair\_iff\_mem\_feasibilitySet}).
\begin{theorem}[\texttt{feasibilitySet\_nonempty\_of\_fair}]
Fair information is consistent: if the truth satisfies every piece of
information then $S\ne\emptyset$.
\end{theorem}
Contrapositively, inconsistency is a certificate that some information is
wrong:
\begin{theorem}[\texttt{exists\_unfair\_of\_feasibilitySet\_eq\_empty}]
If $S=\emptyset$ then for every candidate $h$ there is an $i$ with
$h\notin S_i$.
\end{theorem}

\paragraph{Ideal information.}
The family is \emph{ideal} for $h$ when $S=\{h\}$ (\texttt{STE.Ideal}).
\begin{theorem}[\texttt{Ideal.eq\_of\_mem},
\texttt{ideal\_iff\_fair\_and\_subsingleton}]
Under ideal information every estimate equals the truth; and idealness is
equivalent to fairness together with feasibility being a subsingleton.
\end{theorem}

\paragraph{Topology (Carlson's setting).}
\begin{theorem}[\texttt{isClosed\_feasibilitySet},
\texttt{feasibilitySet\_nonempty\_of\_compact}]
If each $S_i$ is closed then $S$ is closed. If moreover $\Xi$ is compact
and every finite subfamily is consistent (the finite intersection
property, \texttt{FiniteConsistency}), then $S\ne\emptyset$.
\end{theorem}
\noindent The compactness result is the abstract reason STE is well posed
on Carlson's finite key spaces (discrete $\Rightarrow$ compact;
\texttt{isClosed\_feasibilitySet\_of\_discrete}) and on bounded closed
convex constraint sets.

\paragraph{Convexity (Combettes's Hilbert-space setting).}
\begin{theorem}[\texttt{convex\_feasibilitySet}]
If each $S_i$ is convex then $S$ (and every partial feasibility set) is
convex.
\end{theorem}
\noindent This is the geometric precondition for the projection
algorithms of \citet{youla1982,bauschke1996}, a planned extension.

\paragraph{STE decryption skeleton (Carlson).}
A \texttt{CipherSystem} is a keyed family of injective encryption maps.
Observing a ciphertext $c$ induces the key property set
$\{k \mid \exists m\in\mathrm{Meaning},\ \mathrm{enc}\,k\,m=c\}$; the
feasible key set is the STE feasibility set of these.
\begin{theorem}[\texttt{CipherSystem.fair\_of\_genuine},
\texttt{feasibleKeys\_nonempty\_of\_genuine}]
Genuine traffic (ciphertexts that really are encryptions of meaningful
messages under a true key $k_0$) is fair for $k_0$; hence the feasible
key set is nonempty.
\end{theorem}
\begin{theorem}[\texttt{Unicity.eq\_of\_feasible}]
At unicity (feasible key set $=\{k_0\}$, i.e.\ ideal information) every
feasible key is the true key: decryption succeeds by feasibility alone.
\end{theorem}

\section{Verification method}
The environment used to author this work cannot download a Lean toolchain
(egress policy). We therefore treat the GitHub Action as the trusted
verifier: every push to \texttt{main} builds the \texttt{Ste} library
with \texttt{leanprover/lean-action} against the pinned Mathlib cache. A
result is considered established only once its build is green. This
operationalizes the project's rule to distrust results that are not
machine proven.

\section{Future work}
In increasing depth: Carlson's cipher counting (Lemma~4.1, Theorem~4.1);
the reduction of all block ciphers to substitution ciphers (Theorem~5.2,
Corollary~5.3); POCS convergence over Hilbert spaces; and Theorem~2.3,
that property sets are AEP typical sets, placing information theory under
STE. Targets and their open/proven status are tracked in the project
\texttt{log/}.

\bibliographystyle{plainnat}
\bibliography{../../refs}

\end{document}
